Bootstrapping Domain-Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-02-09 paper.tex
edits to introduction

Automatically inducing programs from examples is a long-standing goal
of artificial intelligence. Recent work has successfully used
symbolic search techniques (e.g., Metagol:~\cite{muggleton2015meta},
FlashFill:~\cite{gulwani2011automating}), neural networks trained from
a corpus of examples (e.g., RobustFill:~\cite{devlin2017robustfill},
DeepCoder:~\cite{balog2016deepcoder}), and hybrids of neural and
symbolic methods (e.g., Neural-guided deductive search~\cite{ngds}) to
synthesize programs for task domains such as string transformations,
list processing, and robot navigation and planning. However, all these
approaches -- symbolic, neural and neural-symbolic -- rely upon a
hand-engineered \emph{Domain-Specific Language} (DSL). The DSL is an
inventory of restricted programming primitives, encoding domain-specific
knowledge about the space of programs. In
practice
we are often given only a few examples for each
program to be induced, and thus success often hinges on having a good
DSL, to
provide a crucial inductive bias for what would otherwise be an
unconstrained search through the space of all computable functions.
Here we ask, to what extent can we dispense with such highly
hand-engineered DSLs?

We propose \emph{learning} the DSL. We consider the setting where we
have a collection of related programming tasks, each specified by a
set of input/output examples. We do not assume that the tasks are
annotated with ground-truth programs. We typically will not try to
learn a DSL completely from scratch, but rather to start from a weaker
or more general set of primitives in a Lisp-like representation and
construct a richer, more powerful and better-tuned DSL.

Our algorithm is called \system because it is based on a
novel kind of ``wake-sleep'' learning (c.f. \cite{hinton1995wake}), iterating
between ``wake'' and ``sleep'' phases to achieve three goals: finding
programs that solve each of the tasks; creating a DSL by discovering
and reusing domain-specific subroutines; and training a neural network
that efficiently guides search for programs in the DSL. The DSL
distills commonalities across programs that solve the tasks, letting
the agent solve other related program learning problems. The neural
network ensures that searching for program solutions remains tractable
even as the DSL (and hence the search space for programs) expands.

@@

Viewed as an inference problem,
our goal is to infer the DSL (written $\mathcal{D}$)
along with a program (written $p$) for each set of inputs/outputs (\emph{tasks}, written $x$).
A DSL $\mathcal{D}$ is a collection of domain-specific subroutines.
We equip $\mathcal{D}$ with a real-valued weight vector, $\theta$, and together
$(\mathcal{D},\theta)$ define a distribution over programs.
Fig.~\ref{graphicalModel}(a) diagrams this inference problem as a hierarchical Bayesian model.
Inference in this model is
difficult because the programs are unobserved,
and so we must solve a hard search problem to recover them. To make
search tractable we learn a bottom-up \emph{recognition
  model} (written $q(\cdot )$, Fig.~\ref{graphicalModel}(b)).
The recognition model $q$ is a neural
network that regresses from input/output pairs to a distribution over
programs likely to explain the input/outputs. $q(\cdot )$ implements an amortized
inference
scheme~\cite{le2016inference} for the generative %ritchie2016deep
model.

@@

  \end{tikzpicture}
  \caption{(a) DSL $\mathcal{D}$ generates programs $p$ by sampling DSL primitives with probabilities $\theta$. Observe tasks $x$ (program inputs/outputs). (b)  Neural network $q(\cdot )$, the \emph{recognition model}, regresses from $x$ to the parameters of a distribution over programs,
    $\theta^{(x)}$. Black/red arrows correspond to the generative/recognition models.}\label{graphicalModel}
\end{figure}

To perform inference, we take inspiration from the Helmholtz machine and the Wake/Sleep algorithm~\cite{hinton1995wake},
which alternates between ``waking'', which draws samples from a recognition model (our $q$);
and  ``sleeping'', which trains a recognition model on samples from the generative model (our $\mathcal{D},\theta$).
For us, a wake cycle means using the recognition model to search for programs that solve the tasks.
A sleep cycle means  updating these models so that the DSL and  recognition model
more closely match the distribution of tasks.
An important difference from the Helmholtz machine
is that our setup has two distinct sleep phases: a phase where we update the generative model
$(\mathcal{D},\theta)$, which we call \textbf{Sleep-G},
and a second phase where we update the recognition model $q$,
 called \textbf{Sleep-R}.

@@

We apply \system to four domains:
 building Boolean circuits; symbolic regression; FlashFill-style~\cite{gulwani2011automating} string processing problems; and Lisp-style functions on lists.
 For each of these we deliberately provide an impoverished
 set of programming primitives,
 and show that our algorithm discovers
 its own domain-specific vocabulary for expressing solutions in the domain (Tbl.~\ref{initialExampleDSL}).
 \begin{table}
    \begin{tabular}{ll}
   \toprule
   Domain&Part of the learned DSL\\\midrule
   Circuits&\code{(nand (nand a a) (nand b b))}\\
   &\hspace{0.5cm} \emph{(an OR gate)}\\
   Regression& \code{(+ (* real x) real)} \\
      &\hspace{0.5cm} \emph{(a linear function of x)}\\
   Strings& \code{(join '' (split c s))}\\
   &\hspace{0.5cm} \emph{(delete occurrences of c in string s)}\\
   Lists& \code{(map (lambda (x) (+ x a)) l)}\\
   &\hspace{0.5cm} \emph{(add a to every element of list l)}
   \\\bottomrule
    \end{tabular}
    \caption{ Examples of structure found in DSLs  learned by our algorithm. \system builds a new DSL by discovering and reusing useful subroutines.}\label{initialExampleDSL}
 \end{table}

>>>>

Automatically inducing programs from examples is a long-standing goal
of artificial intelligence. Recent work has successfully used
symbolic search techniques (e.g., Metagol:~\cite{muggleton2015meta},
FlashFill:~\cite{gulwani2011automating}), neural networks trained from
a corpus of examples (e.g., RobustFill:~\cite{devlin2017robustfill},
DeepCoder:~\cite{balog2016deepcoder}), and hybrids of neural and
symbolic methods (e.g., Neural-guided deductive search~\cite{ngds}) to
synthesize programs for task domains such as string transformations,
list processing, and robot navigation and planning. However, all these
approaches -- symbolic, neural and neural-symbolic -- rely upon a
hand-engineered \emph{Domain-Specific Language} (DSL). The DSL is an
inventory of restricted programming primitives, encoding domain-specific
knowledge about the space of programs. In
practice
we are often given only a few examples for each
program to be induced, and thus success often hinges on having a good
DSL to
provide a crucial inductive bias for what would otherwise be an
unconstrained search through the space of all computable functions.
Here we ask, to what extent can we dispense with such highly
hand-engineered DSLs?

We propose \emph{learning} the DSL. We consider the setting where we
have a collection of related programming tasks, each specified by a
set of input/output examples. We do not assume that the tasks are
annotated with ground-truth programs. We typically will not try to
learn a DSL completely from scratch, but rather to start from a weaker
or more general set of primitives and
construct a richer, more powerful, and better-tuned DSL.

Our algorithm is called \system because it is based on a
novel kind of ``wake-sleep'' learning (c.f. \cite{hinton1995wake}), iterating
between ``wake'' and ``sleep'' phases to achieve three goals: finding
programs that solve tasks; creating a DSL by discovering
and reusing domain-specific subroutines; and training a neural network
that efficiently guides search for programs in the DSL. The learned DSL
distills commonalities across programs that solve tasks, helping
the agent solve related program learning problems. The neural
network ensures that searching for program solutions remains tractable
even as the DSL (and hence the search space for programs) expands.

@@

Viewed as an inference problem,
our goal is to infer the DSL (written $\mathcal{D}$)
along with a program (written $p$) for each set of inputs/outputs (\emph{tasks}, written $x$).
A DSL $\mathcal{D}$ is a collection of domain-specific subroutines.
We equip $\mathcal{D}$ with a real-valued weight vector $\theta$, and together
$(\mathcal{D},\theta)$ define a distribution over programs.
Fig.~\ref{graphicalModel}(a) diagrams this inference problem as a hierarchical Bayesian model.
Inference in this model is
difficult because the programs are unobserved,
and so we must solve a hard search problem to recover them. To make
search tractable we learn a bottom-up \emph{recognition
  model} (written $q(\cdot )$, Fig.~\ref{graphicalModel}(b)).
The recognition model $q$ is a neural
network that regresses from input/output pairs to a distribution over
programs likely to explain the input/outputs. $q(\cdot )$ implements an amortized
inference
scheme~\cite{le2016inference} for the generative %ritchie2016deep
model.

@@

  \end{tikzpicture}
  \caption{(a) DSL $\mathcal{D}$ generates programs $p$ by sampling DSL primitives according to probabilities $\theta$. From each program we observe a task $x$ (program inputs/outputs). (b)  Neural network $q(\cdot )$, the \emph{recognition model}, regresses from $x$ to the parameters of a distribution over programs,
    $\theta^{(x)}$. Black/red arrows correspond to the generative/recognition models.}\label{graphicalModel}
\end{figure}

To perform inference, we take inspiration from the Helmholtz machine and the Wake/Sleep algorithm~\cite{hinton1995wake},
which alternates between ``waking'', which draws samples from a recognition model (our $q$);
and  ``sleeping'', which trains a recognition model on samples from the generative model (our $\mathcal{D},\theta$).
For us, a \textbf{Wake} cycle means using the recognition model to search for programs that solve tasks.
A sleep cycle means  updating these models so that the DSL and  recognition model
more closely match the distribution of tasks.
An important difference from the Helmholtz machine
is that our setup has two distinct sleep phases: a phase where we update the generative model
$(\mathcal{D},\theta)$, which we call \textbf{Sleep-G},
and a second phase where we update the recognition model $q$,
 called \textbf{Sleep-R}.

@@

We apply \system to four domains:
 building Boolean circuits; symbolic regression; FlashFill-style~\cite{gulwani2011automating} string processing problems; and Lisp-style functions on lists.
 For each of these we deliberately provide an impoverished
 set of programming primitives,
 and show that our algorithm discovers
 its own domain-specific vocabulary for expressing solutions in the domain (Tbl.~\ref{initialExampleDSL}).
 \begin{table}[b]
    \begin{tabular}{ll}
   \toprule
   Domain&Part of the learned DSL\\\midrule
   Circuits&\code{(nand (nand a a) (nand b b))}\\
   &\hspace{0.5cm} \emph{(an OR gate)}\\
   Regression& \code{(+ (* real x) real)} \\
      &\hspace{0.5cm} \emph{(a linear function of x)}\\
   Strings& \code{(join '' (split c s))}\\
   &\hspace{0.5cm} \emph{(delete occurrences of c in string s)}\\
   Lists& \code{(map (lambda (x) (+ x a)) l)}\\
   &\hspace{0.5cm} \emph{(add a to every element of list l)}
   \\\bottomrule
    \end{tabular}
    \caption{ Examples of structure found in DSLs  learned by our algorithm. \system builds a new DSL by discovering and reusing useful subroutines.}\label{initialExampleDSL}
 \end{table}
